Synthesising Complex Termination Arguments

### 2014-06-03 papers/termination/paper.tex
A whole bunch of restructuring

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}

@@

We propose a general framework that does not assume the existence of linear ranking functions and can uniformly handle:
\begin{itemize}
\item lexicographic linear functions (no templates).
\item non-linear functions.
\item lexicographic non-linear functions.
\end{itemize}

@@

\noindent {\bf Third idea:}
For this purpose, we rephrase the termination problem as a synthesis problem.
 As a direct consequence, the complexity of the method is dependent on the Kolmogorov complexity of the computed ranking function (+ it's not dependent on the size/number of variables of the analysed program). We empirically observe that most programs have ranking functions with low Kolmogorov complexity (see Section ?).

@@

\section{Preliminaries}
\subsection{Termination and Ranking Functions}
A transition system with state space $X$ and transition relation $T \subseteq X \times X$
is said to be \emph{unconditionally terminating} if there is no infinite sequence
$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  We can prove that $T$ is
unconditionally terminating by finding an injective function $R: X \to Y$ where
$Y$ is well-founded and $R$ is monotonically decreasing with respect $T$.  That is
to say:

$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$

\section{Formalism}
\subsection{Syntax and Semantics}

\section{Reducing Termination to Synthesis}
For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
$T$ with initial states I:

\section{Theory}

In the sequel, we will fix the following notation:

@@

@@

@@

\begin{conjecture}
 LKC programs have LKC ranking functions (almost certainly false).
\end{conjecture}

\begin{corollary}
 Goldbach conjecture, Collatz conjecture.
\end{corollary}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}

\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

We propose a general framework that does not assume the existence of linear ranking
functions and can uniformly handle:
\begin{itemize}
\item lexicographic linear functions (no templates).
\item non-linear functions.
\item lexicographic non-linear functions.
\item programs with non-linear operations, e.g. logical and.
\end{itemize}

@@

\noindent {\bf Third idea:}
For this purpose, we rephrase the termination problem as a second-order satisfaction problem
(indeed, a program synthesis problem).
 As a direct consequence, the complexity of the method is dependent on the Kolmogorov
 complexity of the computed ranking function (+ it's not dependent on the size/number
 of variables of the analysed program). We empirically observe that most programs have
 ranking functions with low Kolmogorov complexity (see Section ?).

@@

\section{Preliminaries}
\subsection{Termination and Ranking Functions}
A transition system with state space $X$ and transition relation $T \subseteq X \times X$
is said to be \emph{unconditionally terminating} if there is no infinite sequence
$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  We can prove that $T$ is
unconditionally terminating by finding an injective function $R: X \to Y$ where
$Y$ is well-founded and $R$ is monotonically decreasing with respect $T$.  That is
to say:
$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$

We can specify the existence of a ranking function, and therefore the termination
of a program, using a second order formula:

\begin{definition}[Second-order Termination Formula]
\begin{align*}
 \exists R . \forall x, x' . & R(x) > 0 ~ \wedge \\
                             & T(x, x') \rightarrow R(x) > R(x')
\end{align*}
\end{definition}

A special case of a ranking function is a \emph{linear ranking function}.  This
class of functions is exactly what you'd expect: a linear function that satisfies
the criteria for a ranking function.  We recall that a linear function $f: X \to Y$,
with $\dim(X) = n$ and $\dim(Y) = m$
is one that can be expressed as an $n \times m$ matrix $M$:
$$f(\vec{x}) = M\vec{x}$$

In the case that $\dim(Y) = 1$, this reduces to an inner product:
$$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c$$

If $Y = Z^m$, we say that the ranking function is \emph{lexicographic},
and require that the total order imposed on $Y$ is the lexicographic ordering
induced on tuples of $Z$'s.  We note that some termination arguments
require lexicographic ranking functions, or equivalently, ranking functions
whose co-domain is the ordinals, rather than just $\mathbb{N}$.

\subsection{Machine Arithmetic and Bitvectors}
Computers tend to have fixed width integers.  For a machine with $k$-bit words,
all arithmetic is done modulo $2^k$.  This means that a program's termination
behaviour can depend strongly on whether it is interpreted as operating over
mathematical integers, or over fixed-width machine words.  We can however
observe that if no arithmetic operation \emph{overflows} (i.e. does not result
in a value greater than $2^k$), fixed-width and integer arithmetic coincide.

\section{Formalism}
In this section, we fix a model of computation, describe its semantics and
define the syntax of a language we will work over.

\subsection{Syntax and Semantics}

@@

\section{Combinatorics of Termination}

\begin{itemize}
 \item There are a TON of programs (way more than you'd expect).
 \item There are a TON of terminating programs, and for our model of computation we can count
  how many (the Chaitin constant).
 \item There are a TON of ranking functions (way more than you'd expect, but not many as a
  fraction of programs).
 \item There are not many linear functions.
 \item Most terminating programs don't admit linear ranking functions.
 \item Kolmogorov complexity is relevant.
\end{itemize}

@@

We recall the definition of the Kolmogorov-complexity of a function $f$:

\begin{definition}[Kolmogorov complexity]
 The Kolmogorov complexity $K(f)$ is the length of the shortest program that
 computes $f$.
\end{definition}

We are primarily interested in studying low-Kolmogorov-complexity (LKC)
functions.

\begin{theorem}
 Linear functions are LKC.
\end{theorem}

\begin{proof}
 The program to compute a linear function $f: X \to Y$ is of size linear in
 $\dim(X) \times \dim(Y)$.
\end{proof}

\begin{conjecture}
 Most LKC functions are non-linear.
\end{conjecture}

\begin{corollary}
 LKC is a weaker assumption than linearity.
\end{corollary}

\begin{theorem}
 LKC programs do not always have LKC ranking functions.
\end{theorem}

\begin{proof}
 This would solve the halting problem, Goldbach conjecture, Collatz conjecture.
\end{proof}

\begin{conjecture}
 High-Kolmogorov-complexity (HKC) programs often have LKC ranking functions.
\end{conjecture}

\section{Solving the Synthesis Formula}

\subsection{Reducing Termination to Synthesis}
For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
$T$ with initial states I:

\input{kalashnikov}

\section{Miscellaneous Theory}

@@
